Breakpoint Models Show No Evidence of Recreational Adaptation to Increasing Wildfire Smoke in the American West

### 2021-07-06 MarkdownDocuments/Manuscript_Real.Rmd
added abstract took another pass on intro

---
title: Breakpoint Models Show No Evidence of Recreational Adaptation to Increasing Wildfire Smoke in the American West
author:
- affiliation: Boise State University - Human Environment Systems
  name: Matt Clark
- affiliation: University of Michigan
  name: Alexander Killion
- affiliation: Boise State University
  name: Matthew A. Williamson
- affiliation: Boise State University
  name: Vicken Hillis
- affiliation: Boise State University
  name: Laura? 
date: "`r format(Sys.time(), '%B %d, %Y')`"
keywords: Wildfire, Smoke, Recreation, Park Science
bibliography: [../SmokePaper.bib, ../OtherRefs.bib]
abstract: abstract text
fontsize: 12pt
link-citations: yes
output: 
  pdf_document:
    keep_tex: true
    fig_caption: true
    latex_engine: pdflatex
    template: ~/SmokeProject/MarkdownDocuments/MillerTemplate.tex
header-includes:
- \usepackage{pdfpages}
- \usepackage[font={footnotesize}]{caption}
- \usepackage{bm}
- \usepackage{amsmath}

@@

# Introduction
Humanity must urgently adapt to our increasingly variable and hazardous climate [@IPCC]. Avoiding the worst-case projections of climate change will require adaptations at all levels of human society, from multinational governments to individuals [@IPCC]. Governments and individuals are aware of this, but still are failing to act [@steg_limiting_2018]. Knowledge of adaptive strategies and their benefits does not directly lead to adaptive behavior [@van_valkengoed_meta-analyses_2019]. While governments tend to implement climate adaptation policies in the wake of extreme events, less is known about how individuals are reacting and possibly adapting to a changing climate [@bierbaum_comprehensive_2013; @berrang-ford_tracking_2019; @berrang-ford_are_2011].

A particularly salient environmental effect of climate change in the United States is the increasing wildfire frequency and severity over the last 40 years, primarily in the American West [@kitzberger_contingent_2007; @westerling_increasing_2016]. Because climate change accounts for approximately half of the increase in burned area during this time, fires are expected to worsen over the next 40 years as the climate continues to warm. [@westerling_increasing_2016; @abatzoglou_impact_2016; @yue_ensemble_2013; @wuebbles_climate_2017; @harvey_human-caused_2016; @peterson_new_2003; @barbero_climate_2015]. Accompanying the increase in wildfires in the West, is a projected increase in wildfire smoke [@schroeder_validation_2008]. Fire-prone areas may experience a two-fold increase in wildfire smoke by 2100 [@hurteau_projected_2014].

@@

Wildfire smoke is a complex mixture of gasses aerated by biomass combustion across the landscape [@naeher_woodsmoke_2007]. While many of these compounds have been shown to cause health problems in humans, organic carbon and black carbon particulate matter less than 2.5 $\mu$m in diameter ($\text{PM}_{2.5}$) is particularly harmful to human health [@deryugina_mortality_2019]. $\text{PM}_{2.5}$ Causes between 260,000 - 600,000 global deaths annually and significantly increasing community mortality even under acute exposure [@johnston_fay_h_estimated_2012; @naeher_woodsmoke_2007; @kloog_long-_2013].

Wildfires now account for 50% of all atmospheric $\text{PM}_{2.5}$ in some Western regions compared to less than 20% in 2010 [@burke_changing_2021]. As any level of ambient $\text{PM}_{2.5}$ increases community morbidity, public health agencies increasingly call for individuals to limit outdoor recreational behavior when wildfire smoke is present [@cascio_wildland_2018; @reid_colleen_e_critical_2016; @oneill_summary_2013].

>>>>

---
title: Breakpoint Models Show No Evidence of Recreational Adaptation to Increasing Wildfire Smoke in the American West
author:
- affiliation: Boise State University - Human Environment Systems
  name: Matt Clark
- affiliation: University of Michigan
  name: Alexander Killion
- affiliation: Boise State University
  name: Matthew A. Williamson
- affiliation: Boise State University
  name: Vicken Hillis
- affiliation: Boise State University
  name: Laura? 
date: "`r format(Sys.time(), '%B %d, %Y')`"
keywords: Wildfire, Smoke, Recreation, Park Science
bibliography: [../SmokePaper.bib, ../OtherRefs.bib]
abstract: Humans are going to have to adapt to our increasingly variable and hazardous climate. Rapidly increasing wildfire smoke in the American West represents a serious risk to human health and also a pertinent test-case for individual climate adaptation generally. Here we aggregate data on ambient smoke and recreational visitation in 32 U.S. national parks from 1980 - 2019 to test for observable adaptation to increasing wildfire smoke. We develop a series of breakpoint models to allow for threshold effects in recreational response. These models demonstrate that even when dangerous levels of smoke are present, individuals are failing to limit outdoor recreation. Future research may further our understanding of individual-level climate adaptation by focusing on alternative hazards or finer time series.

fontsize: 12pt
link-citations: yes
output: 
  pdf_document:
    keep_tex: true
    fig_caption: true
    latex_engine: pdflatex
    template: ~/SmokeProject/MarkdownDocuments/MillerTemplate.tex
header-includes:
- \usepackage{pdfpages}
- \usepackage[font={footnotesize}]{caption}
- \usepackage{bm}
- \usepackage{amsmath}

@@

# Introduction
Climate change has already begun to dramatically change the social-ecological landscapes in which humans have learned to operate [@IPCC]. All levels of human society must urgently adapt to these new realities if we are to continue to thrive over the coming millenia [@IPCC]. As a society, we are aware of this, but still are failing to respond adequately [@steg_limiting_2018]. Even when adaptive strategies and their benefits are known, we do not observe adaptive behaviors [@van_valkengoed_meta-analyses_2019]. While research has demonstrated that extreme events tend to spur climate adaptation at the government level, less is known about how individuals are reacting and possibly adapting to a changing climate [@bierbaum_comprehensive_2013; @berrang-ford_tracking_2019; @berrang-ford_are_2011].

A particularly salient environmental effect of climate change in the United States is the increasing wildfire frequency and severity over the last 40 years, primarily in the American West [@kitzberger_contingent_2007; @westerling_increasing_2016]. As climate change accounts for approximately half of the increase in burned area during this time, fires are expected to worsen over the next 40 years as the climate continues to warm [@westerling_increasing_2016; @abatzoglou_impact_2016; @yue_ensemble_2013; @wuebbles_climate_2017; @harvey_human-caused_2016; @peterson_new_2003; @barbero_climate_2015]. Accompanying the increase in wildfires in the American West, is a projected increase in wildfire smoke [@schroeder_validation_2008]. Fire-prone areas may experience a two-fold increase in wildfire smoke by 2100 [@hurteau_projected_2014].

@@

Wildfire smoke is a complex mixture of gasses aerated by biomass combustion across the landscape [@naeher_woodsmoke_2007]. While many of these compounds have been shown to cause health problems in humans, organic carbon and black carbon particulate matter less than 2.5 $\mu$m in diameter ($\text{PM}_{2.5}$) is particularly harmful to human health [@deryugina_mortality_2019]. $\text{PM}_{2.5}$ causes between 260,000 - 600,000 global deaths annually and significantly increases community mortality even under acute exposure [@johnston_fay_h_estimated_2012; @naeher_woodsmoke_2007; @kloog_long-_2013].

In some regions of the American West, wildfires now account for 50% of all atmospheric $\text{PM}_{2.5}$, compared to less than 20% in 2010 [@burke_changing_2021]. As any level of ambient $\text{PM}_{2.5}$ increases community morbidity, public health agencies increasingly call for individuals to limit outdoor recreational behavior when wildfire smoke is present [@cascio_wildland_2018; @reid_colleen_e_critical_2016; @oneill_summary_2013].
